maze-dataset: Maze Generation with Algorithmic Variety and Representational Flexibility

### 2025-04-26 docs/paper/paper.tex
recompile paper

\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{calc}
\tikzset{ % Define a TikZ style for an external hyperlink node.
  hyperlink node url/.style={
    alias=sourcenode,
    append after command={
      let \p1 = (sourcenode.north west),
          \p2 = (sourcenode.south east),
          \n1 = {\x2-\x1},
          \n2 = {\y1-\y2} in
      node[inner sep=0pt, outer sep=0pt, anchor=north west, at=(\p1)]
          {\href{#1}{\XeTeXLinkBox{\phantom{\rule{\n1}{\n2}}}}}
    }
  }
}
\providecommand{\XeTeXLinkBox}[1]{#1}
\newcommand{\docslink}[2]{\href{https://understanding-search.github.io/maze-dataset/#1}{#2}}
\newcommand{\docslinkcode}[2]{\href{https://understanding-search.github.io/maze-dataset/#1}{\texttt{#2}}}
\newcommand{\docslinkcodemain}[1]{\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\##1}{\texttt{#1}}}
\newcommand{\secref}[1]{\hyperref[#1]{section: \textit{\nameref{#1}}}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

  \begin{itemize}
  \tightlist
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}},
    utilizing
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_pixels}{\texttt{as\_pixels()}},
    which can exactly mimic the outputs provided in
    \texttt{easy-to-hard-data}(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild,
    Borgnia, Gupta, Bansal, et al., 2021}) and can be configured to be
    similar to the outputs of Németh
    (\protect\hyperlink{ref-gh_Nemeth_2019}{2019})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_ascii}{\texttt{as\_ascii()}}
    provides a format similar to
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{Oppenheim, 2018};
    \protect\hyperlink{ref-eval-gpt-visual}{Singla, 2023})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/plotting.html\#MazePlot}{\texttt{MazePlot}}
    provides a feature-rich plotting utility with support for multiple
    paths, heatmaps over positions, and more. This is similar to the
    outputs of (\protect\hyperlink{ref-mazegenerator-net}{Alance AB,
    2019}; \protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
    \protect\hyperlink{ref-mathematica-maze}{Guo et al., 2011};
    \protect\hyperlink{ref-mdl-suite}{Nag, 2020})
  \end{itemize}
\item
  The text format provided by
  \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{SolvedMaze(...).as\_tokens()}}
  is similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \&
  Wu, 2023}), but provides over 5.8 million unique formats for
  converting mazes to a text stream, detailed in
  \hyperref[sec:tokenized-output-formats]{section: \textit{\nameref{sec:tokenized-output-formats}}}.
\item
  For rigorous investigations of the response of a model to various
  distributional shifts, preserving metadata about the generation
  algorithm with the dataset itself is essential. To this end, our
  package efficiently stores the dataset along with its metadata in a
  single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images is not only difficult to work with, but also
  inefficient. We use a highly efficient method detailed in
  \hyperref[sec:implementation]{section: \textit{\nameref{sec:implementation}}}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

@@

\begin{figure}
\hypertarget{fig:e2h-raster}{%
\centering
\includegraphics[width=0.3\textwidth,height=\textheight]{figures/maze-raster-input-target.pdf}
\caption{Input is the rasterized maze without the path marked (left),
and provide as a target the maze with all but the correct path removed
(right). Configuration options exist to adjust whether endpoints are
included and if empty cells should be filled in.}\label{fig:e2h-raster}
}
\end{figure}

@@

Autoregressive transformer models can be quite sensitive to the exact
format of input data, and may even use delimiter tokens to perform
reasoning steps (\protect\hyperlink{ref-pfau2024dotbydot}{Pfau et al.,
2024}; \protect\hyperlink{ref-spies2024causalworldmodels}{Spies et al.,
2024}). To facilitate systematic investigation of the effects of
different representations of data on text model performance, we provide
a variety of tokenized text output formats.

We convert mazes to token sequences in two steps. First, the maze is
stringified using
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{as\_tokens()}}.
The
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#MazeTokenizerModular}{\texttt{MazeTokenizerModular}}
class provides a powerful interface for configuring maze stringification
behavior. Second, the sequence of strings is tokenized into integers
using \texttt{encode()}. Tokenization uses a fixed vocabulary for
simplicity. Mazes up to \(50 \times 50\) are supported when using a
unique token for each position, and up to \(128 \times 128\) are
supported when positions in the maze are represented as a pair of
coordinates.

There are many algorithms by which one might tokenize a 2D maze into a
1D format usable by autoregressive text models. Training multiple models
on the encodings output from each of these algorithms may produce very
different internal representations, learned solution algorithms, and
levels of performance. To allow exploration of how different maze
tokenization algorithms affect these models, the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#MazeTokenizerModular}{\texttt{MazeTokenizerModular}}
class contains a rich set of options to customize how mazes are
stringified. This class contains 19 discrete parameters, resulting in
over 5.8 million unique tokenizers. There are 6 additional parameters
available whose functionality is not verified via automated testing, but
further expand the the number of tokenizers by a factor of \(44/3\) to
86 million.

@@

The tokenizer architecture is purposefully designed such that adding and
testing a wide variety of new tokenization algorithms is fast and
minimizes disturbances to functioning code. This is enabled by the
modular architecture and the automatic inclusion of any new tokenizers
in integration tests. To create a new variety of tokenizer, developers
forking the library may simply create their own
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html#_TokenizerElement}{\texttt{\_TokenizerElement}}
subclass and implement the abstract methods. If the behavior change is
sufficiently small, simply adding a parameter to an existing
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html#_TokenizerElement}{\texttt{\_TokenizerElement}}
subclass and updating its implementation will suffice.

The breadth of tokenizers is also easily scaled in the opposite
direction. Due to the exponential scaling of parameter combinations,
adding a small number of new features can significantly slow certain
procedures which rely on constructing all possible tokenizers, such as
integration tests. If any existing subclass contains features which
aren't needed, a developer tool decorator
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization/modular/element_base.html\#mark_as_unsupported}{\texttt{@mark\_as\_unsupported}}
is provided which can be applied to the unneeded
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html#_TokenizerElement}{\texttt{\_TokenizerElement}}
subclasses to prune those features and compact the available space of
tokenizers.

@@

In order to replicate the exact dataset distribution of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}), the parameter
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#MazeDatasetConfig.endpoint_kwargs}{\texttt{MazeDatasetConfig.endpoint\_kwargs:}}
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#EndpointKwargsType}{\texttt{EndpointKwargsType}}
allows for additional constraints such as enforcing that the start or
end point be in a ``dead end'' with only one accessible neighbor cell.
However, combining these constraints with cyclic mazes (such as those
generated with percolation), as was required for the work in
(\protect\hyperlink{ref-knutson2024logicalextrapolation}{Knutson et al.,
2024}), can lead to an absence of valid start and end points. Placing
theoretical bounds on this success rate is difficult, as it depends on
the exact maze generation algorithm and parameters used. To deal with
this, our package provides a way to estimate the success rate of a given
configuration using a symbolic regression model trained with PySR
(\protect\hyperlink{ref-pysr}{Cranmer, 2023}). More details on this can
be found in
\href{https://understanding-search.github.io/maze-dataset/notebooks/estimate_dataset_fractions.html}{\texttt{estimate\_dataset\_fractions.ipynb}}.

@@

We describe mazes with the following representation: for a
\(2\)-dimensional lattice with \(r\) rows and \(c\) columns, we
initialize a boolean array \[
  A = \{0, 1\}^{2 \times r \times c}
\] which we refer to in the code as a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.connection_list}{\texttt{connection\_list}}.
The value at \(A[0,i,j]\) determines whether a \emph{downward}
connection exists from node \([i,j]\) to \([i+1, j]\). Likewise, the
value at \(A[1,i,j]\) determines whether a \emph{rightward} connection
to \([i, j+1]\) exists. Thus, we avoid duplication of data about the
existence of connections and facilitate fast lookup time, at the cost of
requiring additional care with indexing. Note that this setup allows for
a periodic lattice. Generation of mazes is detailed in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMazeGenerators}{\texttt{LatticeMazeGenerators}}.

>>>>

\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{calc}
\tikzset{ % Define a TikZ style for an external hyperlink node.
  hyperlink node url/.style={
    alias=sourcenode,
    append after command={
      let \p1 = (sourcenode.north west),
          \p2 = (sourcenode.south east),
          \n1 = {\x2-\x1},
          \n2 = {\y1-\y2} in
      node[inner sep=0pt, outer sep=0pt, anchor=north west, at=(\p1)]
          {\href{#1}{\XeTeXLinkBox{\phantom{\rule{\n1}{\n2}}}}}
    }
  }
}
\providecommand{\XeTeXLinkBox}[1]{#1}
\newcommand{\docslink}[2]{\href{https://understanding-search.github.io/maze-dataset/#1}{#2}}
\newcommand{\docslinkcode}[2]{\href{https://understanding-search.github.io/maze-dataset/#1}{\texttt{#2}}}
\newcommand{\docslinkcodemain}[1]{\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\##1}{\texttt{#1}}}
\newcommand{\secref}[1]{\hyperref[#1]{section: \textit{\nameref{#1}}}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

  \begin{itemize}
  \tightlist
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}},
    utilizing
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_pixels}{\texttt{as\_pixels()}},
    which can exactly mimic the outputs provided in
    \texttt{easy-to-hard-data}(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild,
    Borgnia, Gupta, Bansal, et al., 2021}) and can be configured to be
    similar to the outputs of Németh
    (\protect\hyperlink{ref-gh_Nemeth_2019}{2019})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_ascii}{\texttt{as\_ascii()}}
    provides a format similar to
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{Oppenheim, 2018};
    \protect\hyperlink{ref-eval-gpt-visual}{Singla, 2023})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/plotting.html\#MazePlot}{\texttt{MazePlot}}
    provides a feature-rich plotting utility with support for multiple
    paths, heatmaps over positions, and more. This is similar to the
    outputs of (\protect\hyperlink{ref-mazegenerator-net}{Alance AB,
    2019}; \protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
    \protect\hyperlink{ref-mathematica-maze}{Guo et al., 2011};
    \protect\hyperlink{ref-mdl-suite}{Nag, 2020})
  \end{itemize}
\item
  The text format provided by
  \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{SolvedMaze(...).as\_tokens()}}
  is similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \&
  Wu, 2023}) but with many more options, detailed
  \hyperref[sec:tokenized-output-formats]{section: \textit{\nameref{sec:tokenized-output-formats}}}.
\item
  Preserving metadata about the generation algorithm with the dataset
  itself is essential for studying the effects of distributional shifts.
  Our package efficiently stores the dataset along with its metadata in
  a single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images or adjacency matrices is not only difficult to
  work with, but also inefficient. We use a highly efficient method
  detailed in
  \hyperref[sec:implementation]{section: \textit{\nameref{sec:implementation}}}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

@@

\input{figures/tex/fig3_raster.tex}

@@

Autoregressive transformer models can be quite sensitive to the exact
format of input data, and may even use delimiter tokens to perform
reasoning steps (\protect\hyperlink{ref-pfau2024dotbydot}{Pfau et al.,
2024}; \protect\hyperlink{ref-spies2024causalworldmodels}{Spies et al.,
2024}). To facilitate systematic investigation of the effects of
different representations of data on text model performance, we provide
a variety of text output formats. By passing an instance of
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#MazeTokenizerModular}{\texttt{MazeTokenizerModular}}
to
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{as\_tokens(...)}},
a maze can be converted to a text sequence. The
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#MazeTokenizerModular}{\texttt{MazeTokenizerModular}}
class contains a rich set of options with class contains 19 discrete
parameters, resulting in over 5.8 million unique possible tokenizers.

@@

@@

In order to replicate the exact dataset distribution of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}), the parameter
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#MazeDatasetConfig.endpoint_kwargs}{\texttt{MazeDatasetConfig.endpoint\_kwargs:}}
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#EndpointKwargsType}{\texttt{EndpointKwargsType}}
allows for additional constraints, such as enforcing that the start or
end point be in a ``dead end'' with only one accessible neighbor cell.
However, combining these constraints with cyclic mazes\footnote{Such as
  those generated with percolation, as was required for the work in
  (\protect\hyperlink{ref-knutson2024logicalextrapolation}{Knutson et
  al., 2024}).} can lead to an absence of valid start and end points. To
deal with this, our package provides a way to estimate the success rate
of a given configuration using a symbolic regression model trained with
PySR (\protect\hyperlink{ref-pysr}{Cranmer, 2023}). More details on this
can be found in
\href{https://understanding-search.github.io/maze-dataset/notebooks/estimate_dataset_fractions.html}{\texttt{estimate\_dataset\_fractions.ipynb}}.

@@

We describe mazes with the following representation: for a
\(2\)-dimensional lattice with \(r\) rows and \(c\) columns, we
initialize a boolean array \[
  A = \{0, 1\}^{2 \times r \times c}
\] which we refer to in the code as a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.connection_list}{\texttt{connection\_list}}.
The value at \(A[0,i,j]\) determines whether a \emph{downward}
connection exists from node \([i,j]\) to \([i+1, j]\). Likewise, the
value at \(A[1,i,j]\) determines whether a \emph{rightward} connection
to \([i, j+1]\) exists. Thus, we avoid duplication of data about the
existence of connections and facilitate fast lookup time, at the cost of
requiring additional care with indexing.
